---
title: "Class 11: AlphaFold"
author: "Sofia Jaravata (A19160915)"
format: pdf
toc: TRUE
---

# Custom Analysis of Resulting Models

```{r}
library(bio3d)
pdb_dir <- read.pdb("hivpr_dimer_0b62a/hivpr_dimer_0b62a_unrelaxed_rank_003_alphafold2_ptm_model_1_seed_000.pdb")

#1. include file directory "hivpr_dimer_23119" FOLDER
#2.  include pdb "model 1" for "file name" 
pdb_dir 
```

Make a vector of input PDB file names that we can read into R

```{r}
pdb_files <- list.files(path = "hivpr_dimer_0b62a/", 
           pattern = "*.pdb", 
           full.names = TRUE)

basename(pdb_files)
```

```{r}
library(bio3d)
# Read all data from Models 
#  and superpose/fit coords

pdbs <- pdbaln(files = pdb_files, fit=TRUE, exefile="msa") 

#if error, CONSOLE: install.packages("BiocManager"), BiocManager::install("msa")
```

```{r}
#Quick view of model sequences 
pdbs
```

> ***RMSD is a standard measure of structural distance between coordinate sets. We can use the `rmsd()` function to calculate the RMSD between all pairs models.*** 

```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

> ***Draw a heatmap of these RMSD matrix values***

```{r}
library(pheatmap)
colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```
Here we can see that models 1 and 3 are more similar to each other than they are to any other model. Models 1 and 2 are quite similar to each other and in turn more similar to model 3 than to models 4 and 5. We will see this trend again in the pLDDT and PAE plots further below.


```{r}
pdb <- read.pdb("1hsg")

plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

> *** `core.find()` function*** 

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```


```{r}
#examine the RMSF between positions of the structure

rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error for domains

```{r}
library(jsonlite)
results_dir <- "hivpr_dimer_0b62a/"
# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
pae_files
```

```{r}
#read the 1st and 5th files

pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

> ***NOTE: The lower the PAE score the better (more reliable model)***

```{r}
pae1$max_pae
pae5$max_pae
```
From this, we can conclude that model 1 is a more reliable model than model 5 as its PAE score is slightly lower. 

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
library(bio3d)
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

***Here is the model 1 plot again but this time using the same data range as the plot for model 5:***

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```
The PAE plots display predicted alignment error between residue pairs. In these plots above, both models appear confident for individual domains but less confident in the relative arrangement between domains, but Model 1 is slightly better due to its lower maximum PAE. 

## Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

How many sequences are in this alignment: 

```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the conserv() function.

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

The conserved Active Site residues will stand out if we generate a consensus sequence with a high cutoff value:

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

For a final visualization of these functionally important sites we can map this conservation score to the Occupancy column of a PDB file for viewing in molecular viewer programs such as Mol*, PyMol, VMD, chimera etc.

```{r}
m1.pdb <- read.pdb(pdb_files[1])

#scaled to fit any number of unique residues  
nres <- length(unique(m1.pdb$atom$resno))

occ <- vec2resno(sim[1:nres], m1.pdb$atom$resno)

write.pdb(m1.pdb, o = occ, file = "m1_conserv.pdb")

```




