---
title: "Class 5: Data Viz with ggplot2"
author: "Sofia Jaravata (A19160915)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**. 

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset: 
```{r}
head(cars)
plot(cars)
```
Before I can use ggplot2 I need to install it on my computer.
To do this, we can use the function `install.packages("ggplot2")`

> *NOTE* We never run `install.packages()` in our quarto doc(We run it once only in R console) as it re-installs package every time we render. 

Once installed we need to load up the package into our R brain :

```{r}
library(ggplot2)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics( how data maps to plot)
- the **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)
The main function in the **ggplot2** package is called `ggplot()`

```{r}
ggplot(cars) + 
  aes(x=speed ,y=dist ) +
  geom_point()
  
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance. 

```{r}
ggplot(cars) + 
  aes(x=speed ,y=dist) +
  geom_point(color = "pink") + 
  geom_smooth(color = "green") +
  labs(title = "Stopping Dist of old cars",
       x = "Speed (MPH)",
       y = "Distance (ft)")
  
```

Q. Can you make the `geom_smooth` function to produce a linear straight line fit to the data and turn off the "gray" error region? 
```{r}
ggplot(cars) + 
  aes(x=speed ,y=dist) +
  geom_point(color = "pink") + 
  geom_smooth(method = lm, se = FALSE, color = "green") +
  labs(title = "Stopping Dist of old cars",
       x = "Speed (MPH)",
       y = "Distance (ft)")
  
```
## Gene Expression Figures

> Import data to plot. 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
sum(genes$State == "up")
```
> A new useful function in this context is the table() function: 

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Modified plot
```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State) +
  geom_point()+
  scale_color_manual(values=c("red", "orange", "yellow"))+
  theme_bw()+
  labs(x="No drug", y= "Drug", 
       title = "Expression changes upon GLP-1 inhibitor treatment")
```

## Going further 

Here we read the famous gapminder dataset:  
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many entries( i.e rows) are in this dataset?

```{r}
nrow(gapminder)
```
 >Q. How many different "country" entries are in this dataset? 
 
```{r}
length(table(gapminder$country))
#OR
length(unique(gapminder$country))
```
 
Let's make our first plot of the entire dataset:
plot of "gdpPercap" vs" lifeExp" colored by "continent":

```{r}
p <- ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point(alpha = 0.3)
p
```
I can add more layers to p, the base plot. 
```{r}
p +
  facet_wrap(~continent)
```


Make a plot for years 1977 and 2007 only(not all years in the dataset).

> Q. First use the **dplyr** package and the `filter()` function from that package to extract the year 2007. 

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
```


```{r}
ggplot(g77)+
  aes(gdpPercap, lifeExp, col=continent, size=pop)+
  geom_point(alpha = 0.5)+
  labs(title = "Gapminder 1977")
```
```{r}
ggplot(g07)+
  aes(gdpPercap, lifeExp, col=continent, size=pop)+
  geom_point(alpha = 0.5)+
  labs(title = "Gapminder 2007")
```

> Q. Make a histogram of lifeExp colored by continent(use `fill=continent` or `col=continent`)
> Q. Make a histogram of lifeExp faceted by continent 

```{r}
ggplot(gapminder)+
  aes(lifeExp, col=continent)+
  geom_histogram()
```
```{r}
ggplot(gapminder)+
  aes(lifeExp, col=continent)+
  geom_histogram() + 
  facet_wrap (~continent)
```

