---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Sofia Jaravata (A19160915)"
format: pdf
toc: true 
---

# Background

Pertussis (more commonly known as **whooping cough**) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis (Figure 1). People of all ages can be infected leading to violent coughing fits followed by a characteristic high-pitched “whoop” like intake of breath. Children have the highest risk for severe complications and death. Recent estimates from the WHO indicate that ~16 million cases and 200,000 infant deaths are due to pertussis annually (Black et al. 2010).

# 1. Investigating pertussis cases by year

### Q1.

> With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called `cdc` and use **ggplot** to make a plot of cases numbers over time.

```{r}
library(datapasta)
cdc <- read.csv("U.S. Reported Pertussis Cases_ 1922 - 2025.csv")
cdc
```

```{r}
cdc$Number.of.Reported.Pertussis.Cases <- as.numeric(gsub("," , "" , cdc$Number.of.Reported.Pertussis.Cases))
```

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(Year, Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x = "Year", 
       y = "Number of cases", 
       title = "Pertussis Cases by Year (1922- )")
```

# 2. A tale of two vaccines (wP & aP)

### Q2. 
> Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(Year, Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=c(1946, 1996),
             linetype = 2, 
             col = c("blue", "red")
             ) + 
  labs(x = "Year", y = "Number of cases", title = "Pertussis Cases by Year (1922- )")

```
I notice that after the introduction of the wP vaccine (blue line), the number of observed cases started to decrease. 

### Q3. 
> Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine in 1996, the number of observed cases started increasing slowly. A possible explanation could be PCR-based testing turning more sensitive, vaccination hesitancy, bacterial evolution, or waning immunity in adolescents with the newer aP vaccine in comparison to the older wP vaccine.  

# 3. Exploring CMI-PB data

## The CMI-PB API returns JSON data

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)

```

### Q4. 
> How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 87 aP and 85 wP infancy vaccinated subjects in the dataset. 

### Q5. 
> How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)

```

There are 112 Female patients and 60 Male patients. 

### Q6. 
> What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
Breakdown of race and biological sex displayed in table above. 

## Side-Note: Working with dates

### Q7. 

> Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
```

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

#Split by aP group, summarize aP ages
ap <- subject %>% filter(infancy_vac == "aP")
round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# Split by wP group, summarize wP ages
wp <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length( wp$age, "years" )))
```

The average age of wP individuals is 37 and 28 for aP individuals. Based on the results, it does seem that they are significantly different as the wP values are greater, and the Median, Mean, 3rd Qu., and Max wP values different and much greater than the aP values. 

### Q8. 
> Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

### Q9. 
> With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

I think these two groups are significantly different as the p-value of 2.372101e-23 indicates that they are different. The aP group has more cases focused in those who are younger with ages ~25-30, while the wP group has more cases spread across all ages, mainly focused in the age range of 30-45, with some cases found in ages 45+. 

## Joining multiple tables

Read the specimen and ab_titer tables into R and store the data as specimen and titer named data frames.

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

### Q9a. 
> Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

### Q10. 
> Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

### Q11. 
> How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

We have **6698** specimens for IgE, **5389** for IgG, **10117** for IgG1 and **10124** for IgG2, IgG3, IgG4. 

### Q12. 
> What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The different $dataset values in abdata are 31520 (2020 dataset), 8085 (2021 dataset), 7301 (2022 dataset), and 5670 (2023 dataset). The number of rows for the most "recent" dataset, 5670 in the 2023 dataset, is much lower than the other datasets, especially in comparison to the number of rows in the 2020 dataset, 31520. 

# 4. Examine IgG Ab titer levels

Now using our joined/merged/linked `abdata` dataset `filter(`) for IgG `isotype`.


```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

### Q13. 
> Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot(show.legend = FALSE) + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

### Q14. 
> What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The FHA and FIM2/3 antigens show differences in the level of IgG antibody titers recognizing them, as well as little changes in PT and PRN, but not as much change in TT, DT, and OVA. This is because these antigens, FHA, FIM2/3, PT, and PRN, are Bordetella pertussis antigens, and they exhibit significantly elevated antibody titers, demonstrating an effective immune response directed at pertussis-specific antigens. OVA is used as a negative control antigen and thus not targeted by the vaccine and TT and DT are more stable comparison antigens. If the immune response is changing after pertussi exposure or vaccination, the antigens showing differences would most likely show it. 

### Q15. 
> Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()+
  labs(title = "OVA antigen levels per visit (aP red, wP teal)")
```

***and the same for antigen=="PT"***

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "PT antigen levels per visit (aP red, wP teal)")
```

### Q16. 
> What do you notice about these two antigens time courses and the PT data in particular?

As time passes, OVA levels stay pretty consistently low among all visits as OVA is the negative control antigen. Over time, PT antibody levels are low overall during visits 1-3, start to increase in visits 4-7, and then decreases to baseline for visits 8-12, indicating that the PT response is variable. This comparison of OVA vs. PT shows that antigen-specific immunity grows against the vaccine. 

### Q17. 
> Do you see any clear difference in aP vs. wP responses?

There are not many clear differences in aP vs. wP responses, they seem pretty similar. Both groups increase in the PT antibody titers, but eventually decrease. Thus, both vaccine types can result in immune responses. 

**Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:**

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

### Q18. 
> Does this trend look similar for the 2020 dataset?

Yes, this trend looks pretty similar for the 2020 dataset. The IgG PT levels are peak around day 14 before their decline. 

# 5. Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

### Q19. 
> Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

### Q20.: 
> What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

I notice that the expression of this gene is at its maximum level right after booster vaccination. The IGHG1 expression is mostly low at many visits, but then spikes early with a maximum level at visit 4 and then randomly later on at visit 8. 

### Q21. 
> Does this pattern in time match the trend of antibody titer data? If not, why not?

No, this pattern in time does not match the trend of antibody titer data because the IGHG1 expression level peaks at visit 4, while the antibody titer data displays variable data and peaks later on. IGHG1 expression measures mRNA inside cells and antibody titer measures secreted antibody protein in blood. IGHG1 expression levels peak earlier and antibody titers peak later because the mRNA must first be translated into proteins before those antibodies are secreted into the bloodstream. 

*We can dig deeper and color and/or facet by infancy_vac status:*

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

There is however no obvious wP vs. aP differences here even if we focus in on a particular visit:

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```




