---
title: 'class12: Lab Session'
author: "Sofia Jaravata (A19160915)"
date: "2026-05-09"
output: pdf_document
toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G in a population 

Downloaded a CSV file from Ensemble < https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel > 

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

## Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)? 

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

**14.0625%**

### GBR
> Now let's look at a different population. I picked the GBR "Great Britain". 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
gbr
```


> Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Let's now dig into this further. 

# Section 4: Population Scale Analysis [HOMEWORK]

## Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

*Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding.*


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

> Total:

```{r}
nrow(expr)
```

> Sample sizes for each: 

```{r}
table(expr$geno)
```

> Median Expression levels: 

```{r}
colnames(expr) 
tapply(expr$exp, expr$geno, median)

```

**There are 108 A|A genotypes, 233 A|G genotypes, and 121 G|G genotypes, thus a total of 462 genotypes. The median expression level is 31.248 for A|A, 25.064 for A|G, and 20.074 for G|G.**

# **Let's make a boxplot**

## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + 
  aes(x=geno, exp, fill = geno) + 
  geom_boxplot(notch=TRUE)
```

**From this boxplot, we can infer that the presence of G|G in this gene correlates to a reduction in gene expression. Thus, the SNP affects ORMDL3 expression, indicating a reduction in its expression.**
